\chapter{Impact and Future Work}

# Impact

## plotly

At the time of writing, **plotly** is the most widely used R package for interactive web graphics according to RStudio's anonymized CRAN mirror download logs. Figure \@ref(fig:cranlogs) shows CRAN downloads for the last year amongst the leading R packages for interactive web graphics. The recent spike in CRAN downloads was due to a major update in **plotly** which introduced a lot of new features.

```{r cranlogs, echo = FALSE, fig.cap = "CRAN downloads over the past year from RStudio's anonymized CRAN mirror download logs. Shown are common packages for interactive web graphics."}
knitr::include_graphics("images/cranlogs")
```

As shown in Figure \@ref(fig:github-stars), **plotly** also enjoys the most GitHub stars among all R packages built on the **htmlwidgets** framework. Github stars provide a mechanism for users of open-source software to indicate their interest in a project on the worlds largest repository of software. Recently, it has started being used to study popularity in open-source projects [@stars].

```{r github-stars, fig.cap = "A screenshot of the htmlwidgets gallery website. This website allows you to browse R packages built on the htmlwidgets framework and sort widgets by the number of GitHub stars."}
knitr::include_graphics("images/github-stars")
```

## LDAvis

Github stars and python port?

# Future work

