\chapter{Impact and Future Work}

# Impact

## plotly

At the time of writing, **plotly** is the most widely used R package for interactive web graphics according to RStudio's anonymized CRAN mirror download logs. When I became the maintainer and [sole contributor](https://github.com/ropensci/plotly/graphs/contributors) of the project over a year ago, several other R packages seemed to be more popular, but as Figure \@ref(fig:cranlogs) suggests, **plotly** is now the most widely used, and by a significant amount. The recent spike in CRAN downloads was due to a [major update](http://moderndata.plot.ly/upgrading-to-plotly-4-0-and-above/) in **plotly** which introduced a lot of new features, but the new linked views and animation framework has yet to even hit CRAN. Once this framework becomes more stable and available on CRAN, I am confident popularity will keep increasing.

```{r cranlogs, echo = FALSE, fig.cap = "CRAN downloads over the past 6 months from RStudio's anonymized CRAN mirror download logs. Shown are common packages for interactive web graphics."}
knitr::include_graphics("images/cranlogs")
```

As shown in Figure \@ref(fig:github-stars), **plotly** also owns the most GitHub stars among all R packages built on the **htmlwidgets** framework. Github stars provide a mechanism for users of open-source software to indicate their interest in a project on the worlds largest repository of software. This metric is even being used in academic studies on the popularity of open-source software projects [@stars].

```{r github-stars, echo = FALSE, fig.cap = "A screenshot of the htmlwidgets gallery website. This website allows you to browse R packages built on the htmlwidgets framework and sort widgets by the number of GitHub stars. As of writing, plotly has 791 stars which is the most among all htmlwidgets."}
knitr::include_graphics("images/github-stars")
```

## LDAvis

The R package **LDAvis** currently has [226 github stars](https://web.archive.org/web/20161122210119/https://github.com/cpsievert/LDAvis) and inspired a python port which currently has [369 github stars](https://web.archive.org/web/20161122210642/https://github.com/bmabey/pyLDAvis). The software was also used to generate a winning submission to [the 2014 statistical graphics video competition](http://web.archive.org/web/20140215133100/http://streaming.stat.iastate.edu/~dicook/video-competition/) and is now featured in the [ASA's video library](http://stat-graphics.org/movies/).

# Future work

I would consider projects mentioned earlier in this thesis (e.g. **pitchRx**, **XML2R**, and **LDAvis**) to be "completed" -- especially from an academic publishing standpoint. The more interesting work to be done involves the projects I've been contributing to over the past 2-3 years -- **animint** and **plotly**. Through this work, I've learned that mixed computation is hard, but it helps tremendously to have collaborators that complement each others strengths. Furthermore, making such a system practically useful requires a _huge_ amount of work that is not directly rewarded with academic publications. In order for such a project to be worthwhile academically (at least within the field of statistical graphics), there needs to a separation of concerns between lower-level rendering details and more novel/experimental extensions. The **animint** project is too ambitious in this sense in that it takes responsibility over everything. On the other hand, the **plotly** project builds upon the plotly.js project, allowing it to leverage the hard work of a thriving open source community and a team of full-time software engineers.

That being said, keeping a project like **plotly** relevant still requires lots of up-keep and maintenance. Aside from things like bug fixes and improving its ability to work well with other software (for example, embedding/linking plots in new computing environments like IPython/Jupyter notebooks, nteract, etc), there are a number of potential projects related to **plotly** that could lead to academic publications:

1. Use what we currently have in **plotly** to assist in deriving insights from novel datasets, like what was done in the chapter [Interactive data visualization on the web using R](#interactive-data-visualization-on-the-web-using-R). 

2. Use what we currently have in **plotly** to study the effects of interactivity on visual perception or design/build novel visualizations to assist in a visual perception task (for example, **LDAvis** could have been created much more quickly with the current infrastructure in **plotly**)

3. Integrate the current infrastructure in **plotly** with **ggplot2** extension packages (i.e., [translate custom ggplot2 geoms](#translating-custom-ggplot2-geoms) to plotly). A particularly intriguing example would be the **ggraph** package -- an extension of ggplot2 aimed at plotting graph-like data structures [@ggraph]. In addition to translating **ggraph** geoms, it would be interesting to explore how plotly's support for [nested selections](#nested-selections) could be integrated to enable linked views between graph structures.

4. Expand plotly's support for linked views without shiny, as eluded to in the [limitations](#limitations) section of [linking views without shiny](#linking-views-without-shiny).
